---
title: "gender_pay_gap_exploratory_simon"
output: html_document
date: "2024-09-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The following document is the documentation of a project work done in the context of the course "Introduction into R" in the Master for statistics in Data Science at the Universidad Carlos III de Madrid.

Topic of this project work will be analyzing the Kaggel Data set https://www.kaggle.com/datasets/nilimajauhari/glassdoor-analyze-gender-pay-gap taken from the job review website Glassdoor. 

This project aims to analyze the gender pay gap using this dataset based 1000 Datapoints consisting out of the Features: 
- Job Title
- Gender
- Age
- Performance evaluation
- Education
- Department
- Seniority
- Base Pay [$/Year]
- Bonus [$/Year]

The goal is to explore the relationships between gender and salary, uncovering correlations between other features and Base Pay/Bonus in the context of gender.

The analysis will begin with a descriptive study of the dataset, where measures of central tendency, variability, and distribution shape for salary data and other continuous variables are calculated. These relationships will be visualized using histograms, box plots, and other graphical representations to assess normality and distribution characteristics.

In the second phase of the project, machine learning methods will be applied to make predictions for Base Pay & Bonus based on the available features using a variety of machine learning libraries like caret and H2O.

## Check Data Contents and Completeness

As an initial step, we review the structure of the diven data by printing the first six rows and checking for missing Values. As can bee seen in the table and the console output, all expected columns are available and now missing values are detected when loading the data from its initial .csv file as a dataframe into the working environment

```{r}
# Load Data
data = data.frame(read.csv("Data/Glassdoor Gender Pay Gap.csv"))
print(head(data))

# check for missing data
total_missing <- sum(is.na(data))
print(paste("Total missing values:", total_missing))

```
Furthermore the kind of values in the individual columns are of interested, which can be further investigated by evaluating the number and types of unique values in each individual column. By doing so, we identify the Columns 

- JobTitle
- Gender
- PerfEval
- Education
- Dept
- Seniority

as being discrete through the limited number of unique values they contain as shown in the first part of the console output. This can be validated by further displaying the specific unique values each of theses features can take as a second step (also shown in the second part of the console output).The remaining Features (Columns):
- Base Pay
- Bonus

vice versa are therefore continuous features where printing each of the respective (992/941) individual values does not make sense. 

These continious features will thus be of greater interest in the next step, where a frequency analysis is performed on all features.

```{r}
print("Number of Unique values")
for (col in colnames(data)) {
  unique_count <- length(unique(data[[col]]))
  print(paste0("Column '", col, "': ", unique_count))
}

paste("\n\n")
print("Unique values for discrete fatures:")
for (col in c("JobTitle","Gender","PerfEval","Education","Dept","Seniority")) {
  unique_values <- unique(data[[col]])
  cat(paste("Column '", col, "': ", paste(unique_values, collapse = ", "), "\n"))
}

```
## Frequency Analysis

To gain a understanding of the distribution of each feature the next step in data discovery is performing a fre


```{r}
par(mar = c(10, 4, 4, 2))

for (col in colnames(data)) {
  if (is.numeric(data[[col]])) {
    # Create histogram for numeric columns
    hist(data[[col]], main = paste("Histogram of", col), xlab = col, col = "lightblue")
  } else {
    # Create bar plot for non-numeric columns
    barplot(table(data[[col]]), main = paste("Bar Plot of", col), xlab = col, col = "lightgreen",las=2)
  }
}
```

## Including Plots


```{r}
# Loop through numeric columns to create overlayed histograms
for (col in colnames(data)) {
  if (is.numeric(data[[col]])) {
    # Set up the plot for the first group
    hist(data[[col]][data$Gender == unique(data$Gender)[1]], 
         main = paste("Overlayed Histogram of", col, "by Gender"),
         xlab = col,
         col = rgb(0, 0, 1, 0.5),  # Blue with transparency
         xlim = range(data[[col]]),  # Set limits based on the data
         breaks = 10)

    # Overlay the histograms for each gender
    for (filter in unique(data$Gender)[-1]) {  # Skip the first element already plotted
      hist(data[[col]][data$Gender == filter], 
           col = rgb(1, 0, 0, 0.5),  # Red with transparency
           add = TRUE, 
           breaks = 10)
    }

    # Add a legend
    legend("topright", 
           legend = unique(data$Gender), 
           fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))
  } else {
    # Bar plot for non-numeric columns
    data_filtered = data[data[[col]] %in% unique(data[[col]]), ]
    barplot(table(data_filtered[[col]]), 
            main = paste("Bar Plot of", col), 
            xlab = col, 
            col = "lightgreen", 
            las = 2)
  }
}
```

